---
title: "Evaluation and explanation"
author: "Alex Zwanenburg"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    includes:
      after_body: license.html
    toc: TRUE
bibliography: "refs.bib"
vignette: >
  %\VignetteIndexEntry{Evaluation and explanation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(familiar)
library(data.table)
```

Familiar will automatically evaluate created models, and export corresponding
tables and plots. The table below lists all available analyses. Several
analyses allow for determining distributions and confidence intervals to provide
an indication of the value spread for a given dataset and models (if
applicable). The `estimation_type` argument defines this behaviour:

-   `point`: Compute point estimates, i.e. single values.

-   `bias_correction` or `bc`: Bias-corrected estimates. A bias-corrected
estimate is computed from (at least) 20 point estimates.

-   `bootstrap_confidence_interval` or `bci`: Bias-corrected estimates with
bootstrap confidence intervals [@Efron2016-ws]. This estimation type allows for
plotting confidence intervals in plots. The number of point estimates required
depends on the `confidence_level` parameter. Familiar uses a rule of thumb of
$n=20/(1-\textrm{confidence_level})$, i.e. at least 400 points for
`confidence_level=0.95`.

Another argument, `detail_level` determines how the required point estimates are
formed:

-   `ensemble`: Point estimates are computed at the ensemble level, i.e. over
all models in the ensemble. This means that, for example, bias-corrected
estimates of model performance are assessed by creating (at least) 20 bootstraps
and computing the model performance of the ensemble model for each bootstrap.

-   `hybrid`: Point estimates are computed from the models in an ensemble.
This means that, for example, bias-corrected estimates of model performance are
directly computed using the models in the ensemble. If there are at least 20
trained models in the ensemble, performance is computed for each model, in
contrast to `ensemble` where performance is computed for the ensemble of models.
If there are less than 20 trained models in the ensemble, bootstraps are created
so that at least 20 point estimates can be made.

-   `model`: Point estimates are computed at the model level. This means that,
for example, bias-corrected estimates of model performance are assessed by
creating (at least) 20 bootstraps and computing the performance of the model for
each bootstrap. Familiar will not automatically create plots in this case.

The more point estimates are required, the more computationally
intensive the analysis is. Some analyses also become computationally more
expensive with more samples. The `sample_limit` argument can be used to specify
the maximum number of samples that should be used.

Some analyses, such as model predictions and individual conditional expectation
plots, do not benefit from bootstraps as they are solely based on values
predicted by the (ensemble) models. If you want to compute bias-corrected
estimates or bootstrap confidence intervals for these analyses you should ensure
that sufficient models are created. Experimental designs such as
`experiment_design="bs(fs+mb,400)+ev"` allow for this, but are computationally
expensive.

| Name                                     | Plot function                                                                     | Export function                       | Estimation type | Detail level | Sample limit |
|------------------------------------------|-----------------------------------------------------------------------------------|---------------------------------------|:---------------:|:------------:|:------------:|
| AUC-PR^a^                                | `plot_auc_precision_recall_curve`                                                 | `export_auc_data`                     |        ×        |      ×       |              |
| AUC-ROC^a^                               | `plot_auc_roc_curve`                                                              | `export_auc_data`                     |        ×        |      ×       |              |
| Calibration info^c^                      |                                                                                   | `export_calibration_info`             |                 |      ×       |              |
| Confusion matrix^a^                      | `plot_confusion_matrix`                                                           | `export_confusion_matrix_data`        |                 |      ×       |              |
| Decision curve analysis^ab^              | `plot_decision_curve`                                                             | `export_decision_curve_analysis_data` |        ×        |      ×       |              |
| Feature expression^c^                    | `plot_sample_clustering`                                                          | `export_feature_expressions`          |                 |              |              |
| Feature selection variable importance^c^ | `plot_feature_selection_occurrence`; `plot_feature_selection_variable_importance` | `export_fs_vimp`                      |                 |              |              |
| Feature similarity^c^                    | `plot_feature_similarity`                                                         | `export_feature_similarity`           |        ×        |              |              |
| Hyperparameters^c^                       |                                                                                   | `export_hyperparameters`              |                 |              |              |
| Individual conditional expectation^c^    | `plot_ice`                                                                        | `export_ice_data`                     |      ×^d^       |      ×       |      ×       |
| Model calibration^ce^                    | `plot_calibration_data`                                                           | `export_calibration_data`             |        ×        |      ×       |              |
| Model performance^c^                     | `plot_model_performance`                                                          | `export_model_performance`            |        ×        |      ×       |              |
| Model predictions^c^                     |                                                                                   | `export_prediction_data`              |      ×^d^       |      ×       |              |
| Model variable importance^ce^            | `plot_model_signature_occurrence`; `plot_model_signature_variable_importance`     | `export_model_vimp`                   |                 |              |              |
| Partial dependence^c^                    | `plot_pd`                                                                         | `export_partial_dependence_data`      |      ×^d^       |      ×       |      ×       |
| Permutation variable importance^c^       | `plot_permutation_variable_importance`                                            | `export_permutation_vimp`             |        ×        |      ×       |              |
| Risk stratification info^b^              |                                                                                   | `export_risk_stratification_info`     |                 |      ×       |              |
| Risk stratification data^b^              | `plot_kaplan_meier`                                                               | `export_risk_stratification_data`     |                 |      ×       |              |
| Sample similarity^c^                     |                                                                                   | `export_sample_similarity`            |        ×        |              |      ×       |
| Univariate analysis^c^                   | `plot_univariate_importance`                                                      | `export_univariate_analysis_data`     |                 |              |              |

: Evaluation and explanation steps.\
^a^ Available for binomial and multinomial outcomes.\
^b^ Available for survival outcomes.\
^c^ Available for all outcomes.\
^d^ Estimation types other than `point` require sufficient models.\
^e^ May not be available for all models.

When familiar is run from `summon_familiar` plots and tables are exported
automatically. The corresponding plot and export functions can all be used externally as well,
as is described in the *Using familiar prospectively* vignette. Indeed, to
showcase some of the functionality we will employ the techniques discussed there.

We will first create two models. We train the first model using (a subset of)
the Wisconsin breast cancer dataset [@Wolberg1990-gn] to predict cell malignancy
from biopsies, which constitutes a classification problem.

```{r, results=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

# Get a dataset.
binomial_data <- familiar:::test.create_good_data_set(outcome_type="binomial")

# Call familiar to create models. Note that we skip hyperparameter optimisation by manually defining the hyperparameters, and that we skip evaluation steps. This is to save time when creating the vignette.
familiar::summon_familiar(data=binomial_data,
                          project_dir=file.path(tempdir()),
                          experiment_dir="binomial",
                          experimental_design="fs+mb",
                          cluster_method="none",
                          fs_method="none",
                          hyperparameter=list("random_forest_ranger"=
                                                list("n_tree"=4,
                                                     "sample_size"=0.7,
                                                     "m_try"=0.5,
                                                     "node_size"=20,
                                                     "tree_depth"=3)),
                          learner="random_forest_ranger",
                          parallel=FALSE,
                          skip_evaluation_elements="all")

# Create path to the directory containing the models.
model_directory_path <- file.path(tempdir(), "binomial", "trained_models", "random_forest_ranger", "none")
model_path <- file.path(model_directory_path, list.files(model_directory_path, pattern="model")[1])

# Load the model.
binomial_model <- readRDS(model_path)
```

The second model is a survival model, and is trained on (a subset of) a colon
chemotherapy clinical trial dataset [@Moertel1995-bh] to predict overall
survival after treatment.

```{r, results=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

# Get a dataset.
survival_data <- familiar:::test.create_good_data_set(outcome_type="survival")

# Call familiar to create models. Note that we skip hyperparameter optimisation by manually defining the hyperparameters, and that we skip evaluation steps. This is to save time when creating the vignette.
familiar::summon_familiar(data=survival_data,
                          project_dir=file.path(tempdir()),
                          experiment_dir="survival",
                          experimental_design="fs+mb",
                          cluster_method="none",
                          fs_method="none",
                          hyperparameter=list("random_forest_ranger"=
                                                list("n_tree"=9,
                                                     "sample_size"=0.95,
                                                     "m_try"=0.68,
                                                     "node_size"=30,
                                                     "tree_depth"=2,
                                                     "alpha"=0.25)),
                          learner="random_forest_ranger",
                          parallel=FALSE,
                          skip_evaluation_elements="all")

# Create path to the directory containing the models.
model_directory_path <- file.path(tempdir(), "survival", "trained_models", "random_forest_ranger", "none")
model_path <- file.path(model_directory_path, list.files(model_directory_path, pattern="model")[1])

# Load the model.
survival_model <- readRDS(model_path)
```

# Model performance

We assess model performance to evaluate accuracy of a model. Model accuracy is
quantified using performance metrics that are documented in the *Performance metrics*
vignette, and can be set using the `evaluation_metric` argument of the
`summon_familiar` function or `metric` for plot and export functions. For
example, this results in the following plot for the breast cancer dataset:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_model_performance(object=binomial_model,
                                          data=binomial_data,
                                          metric = c("auc", "brier", "f1_score"),
                                          x_axis_by="metric")

plots[[1]]
```

The plot shows distributions of the area under the receiver operating
characteristic curve, the brier score and the f1-score. The plot indicates that
the model predicts accurately for the given dataset. This behaviour is expected:
The classification problem is quite easy, and we are evaluating the development
dataset.

To compute model performance metric scores for a dataset, familiar first
predicts expected values for instances in a dataset. Familiar then compares the
expected values against the observed values, and computes the value of the
specified metric.

## Receiver-operating characteristic curve

The receiver-operating characteristic (ROC) curve visualises concordance between
predicted class probabilities and the observed classes [@Hand2001-il]. It shows
the trade-off between sensitivity and specificity of the model. The
receiver-operating characteristic curve for the breast cancer dataset is as
follows:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_auc_roc_curve(object=binomial_model,
                                      data=binomial_data)

plots[[1]]
```

Under ideal circumstances, the curve would lie in the top-left corner of the
plot. If a model does not predict better than at random, the curve would lie
along the diagonal, which corresponds to an AUC-ROC value of 0.50. The plot above
indicates that the model can differentiate between benign and malignant
cells in breast cancer biopsy.

Familiar creates an ROC curve by ascendingly sorting instances by the predicted
probability of the positive class, and computing the number of true positive,
true negative, false positive, false negative cases observed at each instance.
These are then used to compute sensitivity and specificity.

The above plot shows the average of 400 single curves, with 95% confidence
intervals. To achieve this, familiar interrogates each curve over the complete
range of potential specificity values ($[0.000, 1.000]$) with a step size of
$0.005$. Linear interpolation is used for this purpose [@Davis2006-xb]. This
allows for assessing the distribution of sensitivity at fixed specificity
values.

The above approach is the general approach used by familiar. However, familiar
will plot an exact ROC curve if the point estimate (`estimation_type="point"`)
of a single model or the complete ensemble of models is evaluated
(`detail_level="ensemble"`):

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_auc_roc_curve(object=binomial_model,
                                      data=binomial_data,
                                      estimation_type="point")

plots[[1]]
```

Note that we did not specify the detail level to create the above plot because
only a single model is evaluated.

## Precision-recall curve

The precision-recall (PR) curve shows the trade-off between precision (positive
predictive value) and recall (sensitivity). An important distinction between the
PR curve and the ROC curve described above is that precision does not assess
true negatives but false positives. In case the number of negative instances
greatly exceeds the number of positive instances, a large change in the number
of false positives reduces the ROC curve somewhat, but this change is far easier
observed in the PR curve [@Davis2006-xb]. Hence the PR curve is commonly
assessed if the positive class is more important than and rare compared to the
negative class. For the breast cancer dataset, the precision-recall curve is as
follows:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_auc_precision_recall_curve(object=binomial_model,
                                                   data=binomial_data)

plots[[1]]
```

Under ideal circumstances, the curve would lie in the top-right part of the
plot. A random classifier would yield a curve that is mostly horizontal and
located at the fraction of positive instances. This would equal a precision of
45% for the current dataset.

Like the ROC curve, familiar forms the PR curve by ascendingly sorting instances
by the predicted probability of the positive class and computing the number of
true positive and false positive negative cases observed at each instance. These
values are then used to compute recall and precision.

The above plot shows the average of 400 single curves, with 95% confidence
intervals. To achieve this, familiar interrogates each curve over the complete
range of recall values ($[0.000, 1.000]$) with a step size of $0.005$. We use
linear interpolation for this purpose, treating the PR curve as a piecewise
function to manage the situation identified by Davis and Goadrich
[@Davis2006-xb]. Interpolating in this manner allows for assessing the
distribution of precision values at fixed recall values.

Again, this is the general approach used by familiar. Familiar can plot an exact
PR curve when evaluating the point estimate (`estimation_type="point"`) of a
single model or of the complete ensemble of models (`detail_level="ensemble"`).
In the breast cancer dataset, this produces the following curve:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_auc_precision_recall_curve(object=binomial_model,
                                                   data=binomial_data,
                                                   estimation_type="point")

plots[[1]]
```

## Confusion matrix

A confusion matrix schematically represents the number of predicted (expected)
class instances and their actual (observed) values. This may help identify
potential weaknesses of classifiers, i.e. false positives or negatives.

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_confusion_matrix(object=binomial_model,
                                         data=binomial_data)

plots[[1]]
```

Familiar selects the class with the highest predicted class probability as the
expected class. It is currently not possible in familiar to obtain confusion
matrices for set probability thresholds, such as the threshold that maximises
the Matthews correlation coefficient.

## Kaplan-Meier survival curves

Kaplan-Meier survival curves are a standard method for assessing survival over time in
a population, or for assessing survival differences between groups.

Familiar applies one or more thresholds to stratify instances to risk groups.
These thresholds are created during model development to avoid bias. The number
and value of these thresholds is determined by two parameters:
`stratification_method` and `stratification_threshold`. Both parameters are set
when calling `summon_familiar`. `stratification_method` has three options:

-   `median`: Familiar uses the median predicted value in the development cohort
to stratify instances into two risk groups.

-   `fixed`: Instances are stratified based on the sample quantiles of the
predicted values. These quantiles are defined using the
`stratification_threshold` parameter.

-   `optimised`: Use maximally selected rank statistics to determine the optimal
threshold [@Lausen1992-qh; @Hothorn2003-qn] to stratify instances into two
optimally separated risk groups.

The `stratification_threshold` parameter is only used when
`stratification_method="fixed"`. It allows for specifying the sample quantiles
that are used to determine the thresholds. For example
`stratification_threshold=c(0.25,0.75)` creates two thresholds that stratifies
the development dataset into three groups: one with 25% of the instances with
the highest risk, a group with 50% of the instances with medium risk, and a
third group with 25% of instances with lowest risk.

Applying the survival model to the colon cancer dataset yields the following
Kaplan-Meier plot:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_kaplan_meier(object=survival_model,
                                     data=survival_data)

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

In the plot above, familiar divides instances into two groups by the median
threshold in the development dataset. The risk groups are shown with 95%
confidence intervals. Familiar uses the `survival::survfit` function to create
both the survival curves and the confidence intervals. The low-risk group has
significantly better survival than the high-risk group. This is also indicated
in the bottom-left of the main plot, which shows the result of a logrank test
between the two groups. Crosses indicate right-censored patients, of which there
were few. The number of surviving patients in both groups at the time points
along the x-axis are shown below the main plot.

Note that the `predict` method allows for manually specifying stratification
thresholds for existing models.

# Model calibration

Evaluating model performance is important as it allows us to assess how 
accurate a model is. However, it often does not tell us how well we can rely on
predictions for individual instances, e.g. how accurate an instance class
probability is. Model calibration is assessed for this purpose. The model for
the breast cancer dataset produces the following calibration plot:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_calibration_data(object=binomial_model,
                                         data=binomial_data)

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

The calibration curve shows the expected (predicted) probability of malignant
cells versus the observed proportion of malignant cells, together with a 95%
confidence interval. In an ideal calibration plot the calibration curve should
lie on the diagonal (dashed) line. Here the upper part of the curve (expected
probability \> 0.50) follows the diagonal nicely. The lower part of the curve
dips below the diagonal, indicating that predicted probabilities in this range
tend to overestimate malignancy. However, the density plot in the top panel
indicates that predicted probabilities were generally close to 0.0 or 1.0, with
very few instances in between.

Familiar also performs several calibration tests to assess various degrees of
calibration [@Van_Calster2016-ok]. First of all, we assess two aspects of the
linear fit (solid line in the plot): the intercept (calibration-in-the-large) and slope.
These are shown in the top-left of the main plot.

The intercept ideally is 0.00. In our example, it lies below this point, and is
not contained in the 95% confidence interval. We therefore conclude that the
intercept is significantly different at the $\alpha=0.05$ level. Note that this
difference may not be relevant in practice, as it is quite small.

In an ideal case, the slope of the linear fit is 1.00. In our example, the slope
of the linear fit lies above this value. This indicates that, overall, the model
tends to overestimate the probability of malignant cells, but not to a large
degree.

One issue with assessing calibration through a linear fit, is that the
calibration curve can be decidedly non-linear and still produce good values for
the intercept and slope. Therefore, we also assess calibration using statistical
goodness-of-fit tests: the Hosmer-Lemeshow (HL) test for categorical and
regression outcomes [@Hosmer1997-ov], and the Nam-D'Agostino (ND)
[@DAgostino2003-ot] and Greenwood-Nam-D'Agostino (GND) [@Demler2015-yx] tests
for survival outcomes.

## Implementation details

Here we detail the implementation of the model calibration in familiar. The
statistical tests mentioned above, as well as the linear fit, rely on grouping
instances. First we order the instances in the dataset by their expected
(predicted) values. Within each group, the expected values and observed values
are compared:

* Categorical outcomes: We compare the fraction instances with the (positive)
observed class in the group against the mean expected probability for that same
class.

* Regression outcomes: Both expected and observed values are first normalised to
a $[0, 1]$ range using the value range observed in the development dataset. We
then compare the mean observed value in the group against the mean expected
value.

* Survival outcomes: We predict survival probabilities at time points indicated
by `evaluation_times`. The mean expected survival probability in the group is
then compared against the observed probability (Kaplan-Meier estimator) at the
same time points.

We can observe the points thus created in a calibration plot when
`estimation_type="point"`:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_calibration_data(object=binomial_model,
                                         data=binomial_data,
                                         estimation_type="point")

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

The plot with point estimates and the earlier calibration plot with bootstrap
confidence intervals lead to the same assessment. However, you may have observed
that the number of points, and thus the number of groups, is certainly larger
than the 10 groups commonly formed for assessing calibration. This is correct.

Familiar introduces randomness into the assessment of calibration. First,
familiar randomises the number of groups drawn because using a fixed number of
10 groups is arbitrary. Second, the actual group sizes are randomised by drawing
group identifiers with replacement for all instances. We then sort the set of
drawn identifiers, and the instances, ordered by increasing expected value, are
then assigned to their respective group. Thirdly, this process is repeated
multiple times, i.e. 20 times for point estimates, again 20 times (on
bootstraps) for bias-corrected estimates and more for bootstraps confidence
intervals.

Linear fit intercept and slope as well as the goodness-of-fit test p-value are
determined for each single set of groups. Intercept and slopes are then averaged (`point`
and `bias_correction`) or assessed as a distribution
(`bootstrap_confidence_interval`). p-values from goodness-of-fit tests need to
be combined. These tests are not fully independent because they derive
from the same dataset. We therefore compute a harmonic mean p-value
[@Wilson2019-bi] from the individual p-values.

# Decision curve analysis

Decision curve analysis is a method to assess the net (clinical) benefit of
models, introduced by Vickers and Elkin [@Vickers2006-be]. The decision curve
compares the model benefit for different threshold probabilities against two or
more options. The two standard options are that all instances receive an
intervention, and none receive an intervention. What constitutes an intervention
is model-dependent [@Vickers2019-ar]. For our breast cancer cohort, intervention
might constitute treatment of all patients, versus no treatment and treatment for
patients with malignancy predicted by the model. This produces the following
figure:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_decision_curve(object=binomial_model,
                                       data=binomial_data,
                                       y_range=c(-0.2, 0.5))

plots[[1]]
```

The shaded curve is the decision curve produced by the model, with 95% confidence
intervals. The declining curve represents the intervention for all, which offers
no benefit for a malignancy probability of 45%. The horizontal line at a net
benefit of 0.0 represent the no intervention option. As may be observed, the
model offers a net benefit over the entire range of threshold probabilities.

For regression problems no definition of decision curves exists. However,
familiar can perform decision curve analysis for survival outcomes, based on
Vickers et al. [@Vickers2008-uu].

# Variable importance

Not all features are strongly associated with the outcome of interest. One of
the most critical points in machine learning is selecting those features that
are important enough to incorporate into a model. Familiar prospectively
assesses variable importance during the feature selection step using the methods
detailed in the *Feature selection methods* vignette. Later, during evaluation,
familiar employs both model-specific and model-agnostic methods to explaining
which features are considered important by the model itself.

## Model-specific methods

Model-specific methods rely on aspects of the models themselves to determine
variable importance. For instance, the coefficients of regression models are a
measure of variable importance when features are normalised. In the case of the
random forest model for the breast cancer dataset, permutation variable
importance is determined. This produces the following plot:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_model_signature_variable_importance(object=binomial_model,
                                                            data=binomial_data)

plots[[1]]
```

In such plots, we always order features so that the most important features are
on the left, and the least important features on the right.

In case an ensemble of models is assessed, scores are aggregated using an
aggregation method (see the *feature selection methods* vignette). This method
can be set using the `aggregation_method` argument, or the
`eval_aggregation_method` configuration parameter (`summon_familiar`).

The `plot_model_signature_occurrence` method plots the occurrence of features
among the first 5 ranks across an ensemble of models (if available). The rank
threshold can be specified using the `rank_threshold` argument or the
`eval_aggregation_rank_threshold` parameter (`summon_familiar`).

## Permutation variable importance

Permutation variable importance is a model-agnostic variable importance method.
It assesses importance by measuring the decrease in model performance caused by
shuffling values of a feature across the instances in a dataset.

For the breast cancer dataset, this results in the following figure:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_permutation_variable_importance(object=binomial_model,
                                                        data=binomial_data,
                                                        estimation_type="bias_correction")

plots[[1]]
```

The figure shows the features along the y-axis, and the loss in AUC-ROC caused
by permuting the feature values along the x-axis. Accordingly,
`cell_shape_uniformity` is considered to be the most important feature in the
model.

The figure shows two bars for each feature. A well-known issue with permutation
variable importance is that the presence of highly correlated features causes
permutation variable importance to become unreliable [@Hooker2019-pv]. If two
features are both important and highly correlated, shuffling one of them does
not decrease model performance as much. For this reason, familiar will permute
features together depending on their mutual correlation, and the provided
threshold values.

By default, similarity between features is assessed using McFadden's pseudo-$R^2$.
This metric can assess similarity between different types of features, i.e.
numeric and categorical. The default thresholds are 0.3 and 1.0 that
respectively cluster highly similar features and permute all features
individually.

The breast cancer dataset doesn't contain features that exceed the relatively
stringent threshold of 0.3. Since the breast cancer dataset has only
numerical features, we can also measure similarity differently. For example, if we
measure similarity using the Spearman`s rank correlation coefficient, we arrive
at the following:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_permutation_variable_importance(object=binomial_model,
                                                        data=binomial_data,
                                                        estimation_type="bias_correction",
                                                        feature_similarity_metric="spearman",
                                                        feature_similarity_threshold=c(0.6, 1.0))

plots[[1]]
```

This figure paints a different picture: the `cell_shape_uniformity`, `bare_nuclei`,
`normal_nucleoli` and `epithelial_cell_size` features are moderately to strongly
correlated. When permuted together, this has a major effect on model
performance. This can also be observed in the performance decrease caused by
shuffling `cell_shape_uniformity` alone, which only partially matches the
performance decrease caused by shuffling the 4 features. The other 3 features
have little individual impact on performance, yet shuffling them
together directly shows that these features are important as well.

# Partial dependence and individual conditional expectation plots

Variable importance does not explain how features influence the model. Partial
dependence (PD) [@Friedman2001-jf] and individual conditional expectation (ICE)
[@Goldstein2015-rv] plots can be used to visualise this behaviour. As an
example, we can create an ICE plot for the `cell_shape_uniformity` feature that
was found to be important for predicting cell malignancy in the breast cancer.

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_ice(object=binomial_model,
                            data=binomial_data,
                            features="cell_shape_uniformity")

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

The ICE plot shows multiple curves for individual samples, as well as their
average, the partial dependence plot as a somewhat thicker curve. The ICE plot
indicates that an increase in cell shape uniformity is associated by the model
with a rise in malignancy.

It also shows novelty, which is a measure of how out-of-distribution the value
is. Novelty is computed using extended isolation forests [@Hariri2019-xz] and is
the average normalised height at which a sample is found in a terminal leaf node
of the trees in the isolation forest [@Liu2008-kw]. Novelty values range between
`0.00` and `1.00`. Novelty values should be interpreted with care. While increasing
novelty values represent instances that lie further from the distribution, the
values that represent in-distribution instances may vary. In the ICE plot above,
novelty values range between `0.40` and `0.60`. In fact, the median novelty
value for the training instances is `0.47` with an interquartile range of
`0.07`. None of the points in ICE plot can therefore be considered
out-of-distribution.

It is possible to anchor the curves in the ICE plot to a specific feature value.
This allows for assessing the model response for individual instances versus a
fixed value. In effect this can reduce some of the offset caused by other
(fixed) features in each instance, and may help to elucidate the observed
behaviour:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_ice(object=binomial_model,
                            data=binomial_data,
                            features="cell_shape_uniformity",
                            anchor_values=list("cell_shape_uniformity"=10.0))

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

In the plot above, we have anchored the values at `cell_shape_uniformity=10`,
which is the maximum value that this feature can take. All curves are then drawn
relative to the predicted probability found for this particular value. 

It is also possible to show the partial dependence of two features at the same
time:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_ice(object=binomial_model,
                            data=binomial_data,
                            features=c("cell_shape_uniformity", "bare_nuclei"))

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

The partial dependence plot above shows the `cell_shape_uniformity` along the
x-axis and `bare_nuclei` along the y-axis. Higher intensities are associated
with a higher expected probability of malignancy. The figure thus indicates that
higher values of `cell_shape_uniformity` clearly affect the predicted malignancy
probability. Higher values of `bare_nuclei` also lead to higher malignancy
probabilities, but the effect is less pronounced. The average novelty score of
the underlying instances determines the size of each point.

It also possible to anchor 2D partial dependence plots:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_ice(object=binomial_model,
                            data=binomial_data,
                            features=c("cell_shape_uniformity", "bare_nuclei"),
                            anchor_values=list("cell_shape_uniformity"=10.0,
                                               "bare_nuclei"=10.0),
                            show_novelty=FALSE)

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

We disabled novelty (`show_novelty=FALSE`) for plotting the 2D anchored plot.
This changes the appearance of the plot, which now consists of coloured
rectangles. The white dots show the position at which familiar samples the
feature values.

Familiar automatically determines the values at which numerical features are
sampled based on their distribution in the development dataset. Such values can
also be manually set using the `feature_x_range` and `feature_y_range`
arguments. Whereas for the breast cancer dataset the features are linearly
sampled, for the `nodes` feature in the colon cancer dataset this is not the
case:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_ice(object=survival_model,
                            data=survival_data,
                            features=c("nodes", "rx"),
                            evaluation_times=1000,
                            show_novelty=FALSE)

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

Here we see that the number of nodes was not sampled linearly. Survival
probability at 1000 days drops with an increase in nodes. Moreover, the model
predicts that patients in the Levamisole `Lev` and observation `Obs` arms
have lower survival probabilities overall.

# Feature and sample similarity

Highly similar features can carry redundant information for a model. This can be
visualised using similarity heatmaps:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_feature_similarity(object=binomial_model,
                                           data=binomial_data,
                                           feature_similarity_metric="spearman")

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

The heatmap displays Spearman's correlation coefficient between the different
features. The features are also ordered by similarity, so that clusters of
similar features can be identified. Note that the dendrograms depict cluster
distance, i.e. $1-|\rho|$ for Spearman's correlation coefficient.

It is also possible to view the sample clustering. This can be used to identify
if samples are readily stratified into clusters, which may or may not correlate
to the endpoint of interest. For the breast cancer dataset, this yields the
following heatmap:
```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_sample_clustering(object=binomial_model,
                                          data=binomial_data,
                                          feature_similarity_metric="spearman")

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

As may be observed, samples in the breast cancer dataset form two main clusters
that correspond closely to cell malignancy. The top dendrogram is the same as
for feature clustering, as it should be. The right dendrogram is formed after
computing Gower's distance between samples (by default), and shows the division
of samples into two main clusters. The heatmap itself shows normalised values
for the features, based on normalisation and transformation parameters obtained
during model development.

Normalisation options can be altered by changing the `show_normalised_data`
argument. In the breast cancer dataset all features have the same range, so we
can also show the data directly:

```{r, fig.align='center', fig.width=7}
plots <- familiar::plot_sample_clustering(object=binomial_model,
                                          data=binomial_data,
                                          feature_similarity_metric="spearman",
                                          show_normalised_data="none")

grid::grid.newpage()
grid::grid.draw(plots[[1]])
```

# References
